Modeling the near-wake of a vertical-axis cross-flow turbine with 2-D and 3-D RANS

### 2016-03-19 paper.tex
Add introduction from thesis CFD chapter

Despite the development of many simple engineering models based on blade element
momentum or vortex modeling, it remains difficult to predict the performance of
cross-flow turbines in all cases---namely when solidity, or blade chord to
radius ratio is high---a common characteristic of the smaller CFTs. With
computing power becoming evermore available and affordable, computational fluid
dynamics (CFD) presents an attractive method for simulating cross-flow turbines
since it only requires turbulence modeling, i.e., the machine is described from
first principles. However, CFD can be computationally expensive when done in
three dimensions, which may be necessary in some cases. There are many examples
in the literature of 2-D cross-flow turbine simulations with widely varying
results. Balduzzi et al. provides a summary of recent computations and an
attempt to standardize a methodology for using Reynolds-averaged Navier--Stokes
(RANS) to correctly simulate performance of a 2-D CFT \cite{Balduzzi2016}.

Modeling the boundary layer flows on cross-flow turbine blades is very
challenging due to the dynamically changing inflow and angle of attack---which
often exceeds static stall values and causes dynamic stall. Furthermore, the
ability to predict the occurrence and interdependence of boundary layer
transition and separation can have dramatic influence on the blade loading and
therefore the predicted turbine power output. These challenges cast significant
doubt on the ability of high-fidelity CFD modeling to replace wind tunnel or
tank testing of physical models.

\todo[inline]{Put survey of CFD accuracy here}

In this study we set out to model the wake of the University of New Hampshire's
Reference Vertical-Axis (cross-flow) Turbine (UNH-RVAT) using a
Reynolds-averaged Navier--Stokes numerical model. Though studies in the
literature generally focus on predicting the turbine loading and the local blade
boundary layer, we seek to model the larger scale flow produced by the rotor.
The wake of this particular turbine is of interest since it has been shown
experimentally that the near-wake's momentum and energy transport processes are
dominated by vertical advection \cite{Bachant2015-JoT}. It logically follows
that a 2-D simulation would not correctly predict wake recovery, but is of
interest to what extent this is true, since the computational feasibility of 2-D
simulations is attractive.

It is assumed that if the turbine power output and near-wake predictions from
the numerical model match those of the experiments, the flow field can be
inspected in greater detail, i.e., that the experiments will have been
``interpolated''. This will give us access to unmeasured quantities, e.g.
pressure, and allow us to see where the dominant flow structures originate and
help develop low-order wake generator models for use in turbine array modeling.
For example, Shamsoddin and Porte-Agel modeled a 2-D CFT using large eddy
simulation (LES) with an actuator line approach to impart force on the flow
field \cite{Shamsoddin2014}.

To date, little computational work has been done to attempt to design arrays of
CFTs, despite their prospects for closer spacing compared with axial-flow
turbines (AFTs). For example, Araya et al. modeled the flow through a VAT array
using ``leaky Rankine body'' potential flow singularities \cite{Araya2014},
which was able to rank relative performance of array configurations. Goude and
Agren used a 2-D vortex method to simulate a farm of cross-flow turbines
\cite{Goude2010}, though this was not validated with experiments. Durrani et al.
used 2-D CFD to model a group of cross-flow turbines, observing higher power
output for a staggered configuration, but also did not compare with experimental
results \cite{Durrani2011}. Li and Calisal \cite{Li2010} used a 3-D vortex line
method to show mutually improved power output from two adjacent turbines, though
the simulations over-predicted the effects compared with experiments. Antheaume
et al. \cite{Antheaume2008} used a blade element approach coupled with a 3-D
RANS solver to also show how close spacing can improve power output of CFTs.

To summarize, the questions we hope to answer here are:

\begin{enumerate}

    \item Can or should 2-D RANS be used for array design?

@@

    \item Does 3-D RANS realize the correct proportions of wake recovery
    mechanisms, i.e., are the 3-D blade-resolved results a good ``target'' for
    those from a low-order model?

\end{enumerate}

>>>>

Despite the development of many simple engineering models based on blade element
momentum or vortex methods, it remains difficult to predict the performance of
cross-flow turbines (CFTs, often vertical-axis turbines or VATs) in all
cases---namely when solidity, or blade chord-to-radius ratio is high---a common
characteristic of smaller rotors, and those designed for marine hydrokinetic
(MHK) applications. With computing power becoming evermore available and
affordable, computational fluid dynamics (CFD) has become an attractive method
for predicting the performance of cross-flow turbines since it only requires
that turbulence be modeled, i.e., the physics can be described from theoretical
true first principles by solving the Navier--Stokes equations using a boundary
layer resolving body-fitted grid. However, CFD can be computationally expensive
when done in three dimensions, which may be necessary in some cases.

There are many examples in the literature of 2-D cross-flow turbine simulations
with widely varying results. Balduzzi et al.~\cite{Balduzzi2016} provides
a summary of recent efforts and an attempt to standardize a methodology for
using Reynolds-averaged Navier--Stokes (RANS) to correctly predict performance
of a 2-D CFT. Howell et al.~\cite{Howell2010}, performed both 2-D and 3-D
simulations of a high solidity cross-flow turbine using a $k$--$\epsilon$
renormalization group (RNG) turbulence model. The results from the 2-D
simulations over-predicted power coefficient, while the 3-D case matched well
with wind tunnel measurements near the tip speed ratio of maximum power. In
general, 3-D simulations are less common, but have begun to appear more
frequently recently---a testament to the progress towards higher computing
power.

An overview of 3-D blade-resolved cross-flow turbine simulations reported in the
literature is presented in Table~\ref{tab:cfd-refs}. The $k$--$\omega$ SST RANS
turbulence model is shown to be a popular choice due to its success in
predicting flows with adverse pressure gradients and
separation~\cite{Menter2003}. Higher fidelity methods that resolve the large
scales of turbulence, such as large-eddy and detached-eddy simulation have also
been used. Note that for all the studies listed, model validation was either
done for performance or wake predictions---not both---and in some case omitted
entirely.

\begin{table}
    \centering
    \begin{tabular}{c|c|c|c}
        Author & Turbulence modeling & Perf. val. & Wake val. \\ 
        \hline
        Alaimo et al.~\cite{Alaimo2015} & $k$--$\epsilon$ & N/A & N/A \\ 
        Marsh et al.~\cite{Marsh2015} & SST & \cite{Rawlings2008} & N/A \\ 
        Orlandi et al.~\cite{Orlandi2015} & SST & \cite{Akins1989,Mertens2003} & N/A \\ 
        Lam \& Peng~\cite{Lam2016} & SST RANS \& IDDES\tablefootnote{Improved delayed detached-eddy simulation.} & N/A & \cite{Tescione2014} \\ 
        Nini et al.~\cite{Nini2014} & Spalart--Allmaras & N/A & \cite{Battisti2011} \\ 
        Boudreau \& Dumas~\cite{Boudreau2015} & DDES\tablefootnote{Delayed detached-eddy simulation.} & N/A & N/A \\ 
        Li et al.~\cite{Li2013} & SST \& Smagorinsky--Lilly LES & \cite{McLaren2011} & N/A \\ 
        Howell et al.~\cite{Howell2010} & $k$--$\epsilon$ RNG\tablefootnote{Renormalization group.} & \cite{Howell2010} & N/A
    \end{tabular}

    \caption{Selected 3-D blade-resolved cross-flow turbine simulations reported
        in the literature, turbulence modeling employed, and performance and/or wake
        studies used for validation. Note the Li et al. study used periodic boundary
        conditions and is technically considered 2.5-D.}

    \label{tab:cfd-refs}
\end{table}

Modeling the boundary layer flows on cross-flow turbine blades is essential to
predicting the blade loading. This flow condition presents a challenge due to
the dynamically changing inflow velocity and angle of attack---which often
exceeds static stall values and causes dynamic stall. Furthermore, the ability
to predict the occurrence and interdependence of boundary layer transition to
turbulence and separation can have dramatic influence on the blade loading and
therefore the predicted turbine power output. These challenges present
significant obstacles to the prospect of using CFD to replace wind tunnel or
tank testing of physical models.

To date, little computational work has been done to attempt to design arrays of
CFTs, despite their prospects for closer spacing compared with axial-flow
turbines (AFTs). For example, Araya et al.~\cite{Araya2014} modeled the flow
through a VAT array using ``leaky Rankine body'' potential flow singularities,
which was able to rank relative---though not absolut---performance of array
configurations. Goude and Agren~\cite{Goude2010} used a 2-D vortex method to
simulate a farm of cross-flow turbines, though this was not validated with
experiments. Durrani et al.~\cite{Durrani2011} used 2-D CFD to model a
group of cross-flow turbines, observing higher power output for a staggered
configuration, but also did not compare with experimental results. Giorgetti
et al.~\cite{Giorgetti2015} took a similar approach for 2-D array analysis using
turbine pairs inspired by Dabiri~\cite{Dabiri2011}, but again experimental
validation was not performed. Li and Calisal \cite{Li2010} used a 3-D vortex
line method to show mutually improved power output from two adjacent turbines,
though the simulations over-predicted the effects compared with experiments.
Antheaume et al.~\cite{Antheaume2008} used a blade element approach
coupled with a 3-D RANS solver to also show how close spacing can improve power
output of CFTs.

In this study we set out to model the performance and near-wake of the high
solidity University of New Hampshire Reference Vertical Axis Turbine (UNH-RVAT)
using the Reynolds-averaged Navier--Stokes equations discretized via the finite
volume method. Though studies in the literature generally focus on predicting
the turbine loading and the local blade boundary layer, we seek to model both
this and the larger scale flow produced by the rotor, i.e., the near-wake, which
is of interest for this particular turbine since it has been shown
experimentally that the near-wake's momentum and energy transport processes are
dominated by vertical advection \cite{Bachant2015-JoT}. It logically follows
that a 2-D simulation, which omits the vertical dimension, would not correctly
predict wake recovery and turbine--turbine interaction. However, it is of
interest to determine how wrong a 2-D model may be, since the computational
feasibility is attractive.

It is assumed that if the turbine power output and near-wake predictions from
the numerical model match those of the experiments, the flow field can be
inspected in greater detail, i.e., that the experiments will have been
``interpolated.'' This will provide access to unmeasured quantities, e.g.,
pressure, and allow inspection into where the dominant flow structures
originate. This will ultimately help develop and evaluate low-order wake
generator models for use in turbine array modeling. In summary, the questions we
hope to answer here are:

\begin{enumerate}
    \item Can 2-D RANS be used for individual turbine and/or array design?

@@

    \item Does 3-D RANS realize the correct proportions of wake recovery
    mechanisms, i.e., are the 3-D blade-resolved results a good ``target'' for
    those a low-order model should produce?
\end{enumerate}
